
# Install libraries if needed and then load
```{r}
#install.packages(c("tidyverse","dplyr","ggplot2","readxl","writexl","rio","cowplot"))

library(dplyr)
library(tidyverse)
library(readxl)
library(writexl)
library(rio)
library(ggplot2)
library(cowplot)

# Read the state abbreviation list
state_abbrev_dict <- read_excel("state_abbrev_dict.xlsx")
```


# Population dataset cleanup
```{r}
# Read the population by state and year to variable 'population_raw'
population_raw <- read_csv("Population per state per year.csv")
# Remove the first 5 rows with regional data
population_raw <- tail(population_raw, -5) 

# Select the years 2015 to 2019, remove row Puerto Rico 
population_final <-  population_raw %>% 
  select(5, (13:17)) %>%
  subset(!NAME == "Puerto Rico") %>%
  # Left join with 'state_abbrev_dict' 
  left_join(state_abbrev_dict, by = c("NAME" = "State")) %>%
  # Move column Abbr to the first
  relocate(Abbr) %>% 
  # Rename column Abbr to State, NAME to StateName and the rename the years
  rename("state"="Abbr", "statename" ="NAME", "2015" = "POPESTIMATE2015", "2016" = "POPESTIMATE2016", "2017" = "POPESTIMATE2017", "2018" = "POPESTIMATE2018", "2019" = "POPESTIMATE2019") %>%
  # Pivot the data from columns to rows for each year, ignore the State  and StateName columns 
  pivot_longer(cols = !c('state', 'statename'), names_to = "year", values_to = "population") %>%
  # Transform the class of Year column from character to integer
  transform(year = as.integer(year))
```



# PIT count dataset - merging yearly worksheets to one dataframe and cleaning up the data then join with yearly population to calculate homeless rates
```{r}
# Save the file path to variable 'filepath_pit_data'
filepath_pit_data <- "2007-2022-PIT-Counts-by-State.xlsx"

# Using the 'Rio' package import the worksheets in the '2007-2022-PIT-Counts-by-State.xlsx' file into a list 'df_pit_raw'
df_pit_raw <- import_list(filepath_pit_data)

# Create an empty dataframe 'df_pit_count' to store the homeless data
df_pit_count <- data.frame(
  state = character(),
  year = integer(),
  homeless_total = integer(),
  homeless_individual = integer(),
  homeless_family = integer()
)

# Loop through the list 'df_pit_raw' from elements 5 through 9 for the years 2015 to 2019.
for (i in 5:9){
  # Select the columns for State, Overall homeless counts for total, individuals and families and store in 'df_temp'
  df_temp <- select(df_pit_raw[[i]], c('State', 
                                 contains('Overall Homeless,'),
                                 contains('Overall Homeless Individuals,'),
                                 contains('Overall Homeless People in Families,')))
  # Remove the last row that contains the total value
  df_temp <- head(df_temp, -1)
  # Pivot the data from columns to rows for each year, ignore the State column
  df_temp <- df_temp %>% pivot_longer(cols = !'State', 
                                     names_to = c(".value", "year"), 
                                     names_sep = ",")
  # Create a dataframe with the values computed so far
  df_temp <- data.frame(state = df_temp$'State', 
                        year = as.integer(df_temp$year),
                        homeless_total = df_temp$'Overall Homeless',
                        homeless_individual = df_temp$'Overall Homeless Individuals',
                        homeless_family = df_temp$'Overall Homeless People in Families')
  # Bind the 'df_temp' dataframe to the 'df_pit_count' dataframe
  df_pit_count <- rbind(df_pit_count, df_temp)
}

# Remove the observations for "AS", "GU", "MP", "PR", "VI"
df_pit_count <- df_pit_count %>%
  subset(!state %in% c("AS", "GU", "MP", "PR", "VI")) %>%
  arrange(state, year) 

# Do a full join of the population data to the homeless data by state and year
df_pit_count <- df_pit_count %>%
  full_join(population_final, by = c("state" = "state", "year" = "year")) %>%
  # Calculate the rate of homelessness and add them in columns
  mutate(homeless_total_rate = round((homeless_total / population * 100), 2), homeless_individual_rate = round((homeless_individual / population * 100), 2), homeless_family_rate = round((homeless_family / population * 100), 2))
```


# Unemployment dataset cleanup and join to PIT count dataset
```{r}
# Read the unemployment dataset into the variable 'unemployment_raw'
unemployment_raw <- read_excel("Unemployment per state per year.xlsx")

# Pivot the data into the columns Year and unemployment_rate and store in the variable 'unemployment_data_final' 
unemployment_data_final <- tail(unemployment_raw, -1) %>% 
  # Select columns for years 2015 to 2019
  select(1 , (5:9)) %>%
  pivot_longer(cols = !'State', names_to = "year", values_to = "unemployment_rate") %>%
  arrange(State) %>%
  # Transform the class of Year column from character to integer
  transform(year = as.integer(year)) %>%
  rename("statename" = "State")

# Left join 'df_pit_count_final' and 'unemployment_data_final' by state and year into the variable 'df_pit_unemployment'
df_pit_count_final <- df_pit_count %>%
  left_join(unemployment_data_final, by = c("statename" = "statename", "year" = "year"))
```


# Poverty dataset cleanup and join to PIT count dataset
```{r}
# Read the poverty rate dataset into the variable 'poverty_rate'
poverty_rate <- read_excel("Poverty Rates by State 2015-209.xlsx") %>% select(1,2,4)

# Left join 'df_pit_count_final' and 'poverty_rate' by state and year into the variable 'df_pit_count_final'
df_pit_count_final <- df_pit_count_final %>%
  left_join(poverty_rate, by = c("statename" = "state", "year" = "year"))

#sapply(df_pit_count_final, class)
```


# Median income dataset cleanup and join with PIT count dataset
```{r}
median_income <- read_xlsx("Median Household Income 2015-2019.xlsx", col_names = FALSE)
colnames(median_income) <- c("statename", "2019", "2018", "2017", "2016", "2015")
#colnames(median_income)
#sapply(median_income, class)

median_income <- tail(median_income, -2) %>%
  pivot_longer(cols = !'statename', names_to = "year", values_to = "median_income") %>%
  transform(year = as.integer(year))

# Left join 'df_pit_count_final' and 'median_income' by state and year into the variable 'df_pit_count_final'
df_pit_count_final <- df_pit_count_final %>%
  left_join(median_income, by = c("statename" = "statename", "year" = "year"))
```


# FMR dataset cleanup and join with PIT count dataset
```{r}
# Read FMR data from 2015 to variable fmr_2015
fmr_2015 <- read_excel("FY2015F_4050_Final.xls") %>%
  # Select fmr for studio, 1BHK, 2BHK apartments and the state
   select(fmr0,fmr1, fmr2, state = state_alpha) %>%
   group_by(state) %>%
   # Group by state and calculate the mean for FMR for studio, 1 and 2 BHK apartment
   # Round up or down to nearest whole number
   summarise(fmr_studio = round(mean(fmr0)),fmr_1br = round(mean(fmr1)), fmr_2br = round(mean(fmr2))) %>%
   ungroup() %>%
   # Remove US territories 
   subset(!state %in% c("AS", "GU", "MP", "PR", "VI")) %>% 
   # Add the year to the table
   add_column(year = "2015")


# Read FMR data from 2015 to variable fmr_2016 and repeat steps 
fmr_2016 <- read_excel("FY2016F_4050_Final.xlsx") %>%
  select(fmr0,fmr1, fmr2, state = state_alpha) %>%
  group_by(state) %>%
  summarise(fmr_studio = round(mean(fmr0)),fmr_1br = round(mean(fmr1)), fmr_2br = round(mean(fmr2))) %>%
  ungroup() %>%
  subset(!state %in% c("AS", "GU", "MP", "PR", "VI")) %>% 
  add_column(year = "2016")

# Read FMR data from 2017 to variable fmr_2017 and repeat steps 
fmr_2017 <- read_excel("FY2017-4050-County-Level_Data.xlsx") %>%
 select(fmr0,fmr1, fmr2, state = state_alpha) %>%
  group_by(state) %>%
  summarise(fmr_studio = round(mean(fmr0)),fmr_1br = round(mean(fmr1)), fmr_2br = round(mean(fmr2))) %>%
  ungroup() %>%
  subset(!state %in% c("AS", "GU", "MP", "PR", "VI")) %>% 
  add_column(year = "2017")

# Read FMR data from 2018 to variable fmr_2018 and repeat steps 
fmr_2018 <- read_excel("FY18_4050_FMRs_rev.xlsx") %>%
  select(fmr_0,fmr_1, fmr_2, state = state_alpha) %>%
  group_by(state) %>%
  summarise(fmr_studio = round(mean(fmr_0)),fmr_1br = round(mean(fmr_1)), fmr_2br = round(mean(fmr_2))) %>%
  ungroup() %>%
  subset(!state %in% c("AS", "GU", "MP", "PR", "VI")) %>% 
  add_column(year = "2018")

# Read FMR data from 2019 to variable fmr_2019 and repeat steps 
fmr_2019 <- read_excel("FY2019_4050_FMRs_rev2.xlsx") %>%
  select(fmr_0, fmr_1, fmr_2, state = state_alpha) %>%
  group_by(state) %>%
  summarise(fmr_studio = round(mean(fmr_0)),fmr_1br = round(mean(fmr_1)), fmr_2br = round(mean(fmr_2))) %>%
  ungroup() %>%
  subset(!state %in% c("AS", "GU", "MP", "PR", "VI")) %>% 
  add_column(year = "2019")

# Bind the data from 2015 to 2019 into one table and convert year to integer
fmr_2015_19 <- bind_rows(fmr_2015, fmr_2016,fmr_2017, fmr_2018, fmr_2019) %>%
  transform(year = as.integer(year))

# Left join 'df_pit_count_final' and 'fmr_2015_19' by state and year into the variable 'df_pit_count_final'
df_pit_count_final <- df_pit_count_final %>%
  left_join(fmr_2015_19, by = c("state" = "state", "year" = "year"))
```


# Gross Rent dataset cleanup and join with PIT count dataset
```{r}
# Read the file
gross_rent_raw <- read_excel("Median Gross Rent by Bedrooms 2015-2019.xlsx")
# Remove the last row that has total
gross_rent_raw <- head(gross_rent_raw, -1)




# Create a table for 2015 with coloumns for 2015
gross_rent_2015 <- gross_rent_raw %>%
  select(1, (3:5))
# Change the column names
colnames(gross_rent_2015) <- c("statename", "mgr_studio", "mgr_1br", "mgr_2br")
# Add the year
gross_rent_2015 <- gross_rent_2015 %>% add_column(year = "2015")

# Repeat same spets for 2016
gross_rent_2016 <- gross_rent_raw %>%
  select(1, (7:9))
colnames(gross_rent_2016) <- c("statename", "mgr_studio", "mgr_1br", "mgr_2br")
gross_rent_2016 <- gross_rent_2016 %>% add_column(year = "2016")

# Repeat same spets for 2017
gross_rent_2017 <- gross_rent_raw %>%
  select(1, (11:13))
colnames(gross_rent_2017) <- c("statename", "mgr_studio", "mgr_1br", "mgr_2br")
gross_rent_2017 <- gross_rent_2017 %>% add_column(year = "2017")

# Repeat same spets for 2018
gross_rent_2018 <- gross_rent_raw %>%
  select(1, (15:17))
colnames(gross_rent_2018) <- c("statename", "mgr_studio", "mgr_1br", "mgr_2br")
gross_rent_2018 <- gross_rent_2018 %>% add_column(year = "2018")

# Repeat same spets for 2019
gross_rent_2019 <- gross_rent_raw %>%
  select(1, (19:21))
colnames(gross_rent_2019) <- c("statename", "mgr_studio", "mgr_1br", "mgr_2br")
gross_rent_2019 <- gross_rent_2019 %>% add_column(year = "2019")

# Bind the data from 2015 to 2019 into one table and convert year to integer
mgr_2015_19 <- bind_rows(gross_rent_2015, gross_rent_2016,gross_rent_2017, gross_rent_2018, gross_rent_2019) %>%
  transform(year = as.integer(year))

# Left join 'df_pit_count_final' and 'mgr_2015_19' by state and year into the variable 'df_pit_count_final'
df_pit_count_final <- df_pit_count_final %>%
  left_join(mgr_2015_19, by = c("statename" = "statename", "year" = "year"))

```


# Minimum wage dataset cleanup and join with PIT count dataset
```{r}
# Read dat from csv file and select relevant columns
min_wage <- read_csv("Minimum Wage Data.csv") %>%
  select(1:3) %>% 
  # Filter the years 2015 to 2019, remove US territories, rename columns
  filter(Year %in% c("2015", "2016", "2017", "2018", "2019") & !State %in% c("U.S. Virgin Islands", "Guam", "Puerto Rico", "PR", "VI")) %>%
  rename("year" = "Year", "statename" = "State", "min_wage" = "State.Minimum.Wage")

# Left join 'df_pit_count_final' and 'min_wage' by state and year into the variable 'df_pit_count_final'
df_pit_count_final <- df_pit_count_final %>%
  left_join(min_wage, by = c("statename" = "statename", "year" = "year"))

# Convert all states in column 'statename' to lowercase 
df_pit_count_final$statename <- tolower(df_pit_count_final$statename)
```